Simplified-spatial-availability-example

### 2022-05-27 Simplified-spatial-availability-example.Rmd
Final audit of the procedures by AS and AP

Plot:
```{r}
rbind(x_transit,
      x_car_motor,
      x_bike,
      x_walk) %>%
  mutate(mode = factor(mode,
                       levels = c("bike", "walk", "car/motor", "transit"))) %>%
  ggplot(aes(x = t, y = f)) + 
  geom_line() + 
  facet_wrap(~ mode, nrow = 2)
```

What do the impedance curves tell us about the travel behavior by different modes? Check the histograms:
```{r}
rbind(transit_HW_alltt,
      car_motor_HW_alltt,
      bike_HW_alltt,
      walk_HW_alltt) %>%
  mutate(mode = factor(mode,
                       levels = c("bike", "walk", "car/motor", "transit"))) %>%
  ggplot(aes(x = TIEMPO_VIAJES, weights = ELE_G_POND_ESC2, after_stat(density))) + 
  geom_histogram(bins = 35) + 
  facet_wrap(~ mode, nrow = 2)
```

@@

# Instead of creating various columns like above, create a single column with the mode
OD_v_HW <- OD_v_HW  %>% 
  mutate(mode = case_when(MODO_PRIORITARIO <= 9 ~ "transit",
                          MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10 ~ "car_motor",
                          MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20 ~ "bike",
                          MODO_PRIORITARIO == 24 ~ "walk"),
         mode = factor(mode))

@@

NOW WE NEED: the travel times from the routing algorithm between origin-destination pairs. Anastasia did not use the routing, it may not be needed; instead she used the average travel time by mode for all trips

<!--
Next join workers and jobs to the ZN simple features object:
```{r}
load("data/zn.RData")

@@

Then add the number of workers and jobs from each ZN to each trip:
```{r}
#add the number of jobs and workers to the od_ft_tt matrix
OD <- OD_v_HW %>% merge(zn %>% dplyr::select(ZT1259, workers_transit,workers_car,workers_bike,workers_walk) %>% st_drop_geometry(),
                        by.x = "VORIZT1259", by.y="ZT1259", all.x = TRUE)

OD <- OD %>% merge(zn %>% dplyr::select(ZT1259, jobs_transit,jobs_car,jobs_bike,jobs_walk) %>% st_drop_geometry(),
                   by.x = "VDESZT1259", by.y="ZT1259", all.x = TRUE)

OD <- OD %>% 
  mutate(workers_transit = ifelse(is.na(workers_transit), 0, workers_transit),
         workers_car = ifelse(is.na(workers_car), 0, workers_car),
         workers_bike = ifelse(is.na(workers_bike), 0, workers_bike),
         workers_walk = ifelse(is.na(workers_walk), 0, workers_walk),
         jobs_car = ifelse(is.na(jobs_car), 0, jobs_car),
         jobs_transit = ifelse(is.na(jobs_transit), 0, jobs_transit),
         jobs_bike = ifelse(is.na(jobs_bike), 0, jobs_bike),
         jobs_walk = ifelse(is.na(jobs_walk), 0, jobs_walk))
```

@@

```{r}
OD_test1 <- OD %>%
  group_by(VORIZT1259,VDESZT1259) %>%
  summarise(trips_transit = sum(trips_transit),
            trips_car_motor = sum(trips_car_motor),
            trips_bike = sum(trips_bike),
            trips_walk = sum(trips_walk),
            workers_transit = mean(workers_transit), #workers_ and jobs_ values are retained from the ZN object - so they are the same value for each OD pair thus we use mean() here.
            workers_car = mean(workers_car),
            workers_bike = mean(workers_bike),
            workers_walk = mean(workers_walk),
            jobs_transit = mean(jobs_transit),
            jobs_car = mean(jobs_car),
            jobs_bike = mean(jobs_bike),
            jobs_walk = mean(jobs_walk)) %>% ungroup()

OD_test_transit <- OD %>% filter(trips_transit == 1) %>%
  group_by(VORIZT1259,VDESZT1259) %>%
  summarise(tt_transit = mean(TIEMPO_VIAJES, na.rm=T)) %>% ungroup()

OD_test_car <- OD %>% filter(trips_car_motor == 1) %>%
  group_by(VORIZT1259,VDESZT1259) %>%
  summarise(tt_car = mean(TIEMPO_VIAJES, na.rm=T)) %>% ungroup()

OD_test_bike <- OD %>% filter(trips_bike == 1) %>%
  group_by(VORIZT1259,VDESZT1259) %>%
  summarise(tt_bike = mean(TIEMPO_VIAJES)) %>% ungroup()

OD_test_walk <- OD %>% filter(trips_walk == 1) %>%
  group_by(VORIZT1259,VDESZT1259) %>%
  summarise(tt_walk = mean(TIEMPO_VIAJES)) %>% ungroup()

OD_test <- OD_test1 %>% merge(OD_test_transit, by=c("VORIZT1259","VDESZT1259"), all=T)
OD_test <- OD_test %>% merge(OD_test_car, by=c("VORIZT1259","VDESZT1259"), all=T)
OD_test <- OD_test %>% merge(OD_test_bike, by=c("VORIZT1259","VDESZT1259"), all=T)
OD_test <- OD_test %>% merge(OD_test_walk, by=c("VORIZT1259","VDESZT1259"), all=T)

#check, should add up to the origin number of trips 27795 - perfect!
sum(OD$trips_transit) + sum(OD$trips_car_motor) + sum(OD$trips_bike) + sum(OD$trips_walk)
sum(OD_test$trips_transit) + sum(OD_test$trips_car_motor) + sum(OD_test$trips_bike) + sum(OD_test$trips_walk)
```

Check a sample OD pair:
```{r}
OD_test %>% filter(VORIZT1259 == "004-001" & VDESZT1259 == "004-001")
OD %>% filter(VORIZT1259 == "004-001" & VDESZT1259 == "004-001" & trips_car_motor == "1" )
```
Good the averages make sense.

now estimate f value using these averaged travel times:
```{r, echo=FALSE, warning=FALSE, message=FALSE}
#dgamma is the density plot for gamma function and dlnorm is the desntiy plot for the lognormal function; adding values as "f" our impedance function
OD_test <- OD_test %>%
  mutate(f_car = dgamma(tt_car, HW_car_motor_gamma$estimate["shape"], HW_car_motor_gamma$estimate["rate"]),
         f_car = ifelse(is.na(f_car), 0, f_car),
         f_transit = dlnorm(tt_transit, HW_transit_lnorm$estimate["meanlog"], HW_transit_lnorm$estimate["sdlog"]),
         f_transit = ifelse(is.na(f_transit), 0, f_transit),
         f_bike = dlnorm(tt_bike, HW_bike_lnorm$estimate["meanlog"], HW_bike_lnorm$estimate["sdlog"]),
         f_bike = ifelse(is.na(f_bike), 0, f_bike),
         f_walk = dlnorm(tt_walk, HW_walk_lnorm$estimate["meanlog"], HW_walk_lnorm$estimate["sdlog"]),
         f_walk = ifelse(is.na(f_walk), 0, f_walk))
```

>>>>

Bind the empirical and theoretical trip length distributions:
```{r}
tld_empirical <- rbind(transit_HW_alltt,
      car_motor_HW_alltt,
      bike_HW_alltt,
      walk_HW_alltt) %>%
  mutate(mode = factor(mode,
                       levels = c("bike", "walk", "car/motor", "transit")),
         distribution = "empirical")

tld_theoretical <- rbind(x_transit,
      x_car_motor,
      x_bike,
      x_walk) %>%
  mutate(mode = factor(mode,
                       levels = c("bike", "walk", "car/motor", "transit")))
```

Plot:
```{r}
ggplot() + 
  geom_histogram(data = tld_empirical,
                 aes(x = TIEMPO_VIAJES, 
                     #weights = ELE_G_POND_ESC2, 
                     after_stat(density)),
                 bins = 45) +
  geom_line(data = tld_theoretical,
            aes(x = t, y = f)) + 
  facet_wrap(~ mode, nrow = 2)
```

Double check the empirical trip length distributions:
```{r}
rbind(transit_HW_alltt,
      car_motor_HW_alltt,
      bike_HW_alltt,
      walk_HW_alltt) %>%
  mutate(mode = factor(mode,
                       levels = c("bike", "walk", "car/motor", "transit"))) %>%
  ggplot(aes(x = TIEMPO_VIAJES, weights = ELE_G_POND_ESC2, after_stat(density))) + 
  geom_histogram(bins = 45) + 
  facet_wrap(~ mode, nrow = 2)
```

@@

# Instead of creating various columns like above, create a single column with the mode
OD_v_HW <- OD_v_HW  %>% 
  mutate(mode = case_when(MODO_PRIORITARIO <= 9 ~ "transit",
                          MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10 ~ "car/motor",
                          MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20 ~ "bike",
                          MODO_PRIORITARIO == 24 ~ "walk"),
         mode = factor(mode,
                       levels = c("bike", "walk", "car/motor", "transit")))

@@

Next join workers and jobs to the ZN simple features object:
```{r}
load("data/zn.RData")

@@

Then add the number of workers and jobs from each ZN to each trip:
```{r}
#add the number of jobs and workers to the od_ft_tt matrix
# OD <- OD_v_HW %>% merge(zn %>% dplyr::select(ZT1259, workers_transit,workers_car,workers_bike,workers_walk) %>% st_drop_geometry(),
#                         by.x = "VORIZT1259", by.y="ZT1259", all.x = TRUE)
# 
# OD <- OD %>% merge(zn %>% dplyr::select(ZT1259, jobs_transit,jobs_car,jobs_bike,jobs_walk) %>% st_drop_geometry(),
#                    by.x = "VDESZT1259", by.y="ZT1259", all.x = TRUE)
# 
# OD <- OD %>% 
#   mutate(workers_transit = ifelse(is.na(workers_transit), 0, workers_transit),
#          workers_car = ifelse(is.na(workers_car), 0, workers_car),
#          workers_bike = ifelse(is.na(workers_bike), 0, workers_bike),
#          workers_walk = ifelse(is.na(workers_walk), 0, workers_walk),
#          jobs_car = ifelse(is.na(jobs_car), 0, jobs_car),
#          jobs_transit = ifelse(is.na(jobs_transit), 0, jobs_transit),
#          jobs_bike = ifelse(is.na(jobs_bike), 0, jobs_bike),
#          jobs_walk = ifelse(is.na(jobs_walk), 0, jobs_walk))
```

@@

```{r}
# OD_test1 <- OD %>%
#   group_by(VORIZT1259,VDESZT1259) %>%
#   summarise(trips_transit = sum(trips_transit),
#             trips_car_motor = sum(trips_car_motor),
#             trips_bike = sum(trips_bike),
#             trips_walk = sum(trips_walk),
#             workers_transit = mean(workers_transit), #workers_ and jobs_ values are retained from the ZN object - so they are the same value for each OD pair thus we use mean() here.
#             workers_car = mean(workers_car),
#             workers_bike = mean(workers_bike),
#             workers_walk = mean(workers_walk),
#             jobs_transit = mean(jobs_transit),
#             jobs_car = mean(jobs_car),
#             jobs_bike = mean(jobs_bike),
#             jobs_walk = mean(jobs_walk)) %>% ungroup()

OD <- OD_v_HW %>%
  group_by(VORIZT1259, 
           VDESZT1259,
           mode) %>%
  summarize(tt = mean(TIEMPO_VIAJES),
            .groups = "drop")

# Check the mean and median travel times by mode
OD %>% 
  group_by(mode) %>% 
  summarize(mtt = mean(tt))
OD %>% 
  group_by(mode) %>% 
  summarize(mtt = median(tt))

# OD_test_transit <- OD %>% filter(trips_transit == 1) %>%
#   group_by(VORIZT1259,VDESZT1259) %>%
#   summarise(tt_transit = mean(TIEMPO_VIAJES, na.rm=T)) %>% ungroup()
# 
# OD_test_car <- OD %>% filter(trips_car_motor == 1) %>%
#   group_by(VORIZT1259,VDESZT1259) %>%
#   summarise(tt_car = mean(TIEMPO_VIAJES, na.rm=T)) %>% ungroup()
# 
# OD_test_bike <- OD %>% filter(trips_bike == 1) %>%
#   group_by(VORIZT1259,VDESZT1259) %>%
#   summarise(tt_bike = mean(TIEMPO_VIAJES)) %>% ungroup()
# 
# OD_test_walk <- OD %>% filter(trips_walk == 1) %>%
#   group_by(VORIZT1259,VDESZT1259) %>%
#   summarise(tt_walk = mean(TIEMPO_VIAJES)) %>% ungroup()
# 
# OD_test <- OD_test1 %>% merge(OD_test_transit, by=c("VORIZT1259","VDESZT1259"), all=T)
# OD_test <- OD_test %>% merge(OD_test_car, by=c("VORIZT1259","VDESZT1259"), all=T)
# OD_test <- OD_test %>% merge(OD_test_bike, by=c("VORIZT1259","VDESZT1259"), all=T)
# OD_test <- OD_test %>% merge(OD_test_walk, by=c("VORIZT1259","VDESZT1259"), all=T)
# 
# #check, should add up to the origin number of trips 27795 - perfect!
# sum(OD$trips_transit) + sum(OD$trips_car_motor) + sum(OD$trips_bike) + sum(OD$trips_walk)
# sum(OD_test$trips_transit) + sum(OD_test$trips_car_motor) + sum(OD_test$trips_bike) + sum(OD_test$trips_walk)
```

<!--

Check a sample OD pair:
```{r}
OD_test %>% filter(VORIZT1259 == "004-001" & VDESZT1259 == "004-001")
OD %>% filter(VORIZT1259 == "004-001" & VDESZT1259 == "004-001" & trips_car_motor == "1" )
```
Good the averages make sense.
-->

Join workers and jobs to OD table:
```{r}
OD_2 <- OD %>%
  left_join(workers,
            by = c("VORIZT1259" = "ZT1259", "mode"))

OD_2 <- OD_2 %>%
  left_join(jobs,
            by = c("VDESZT1259" = "ZT1259"))
```

Now estimate f value using these averaged travel times:
```{r, echo=FALSE, warning=FALSE, message=FALSE}
#dgamma is the density plot for gamma function and dlnorm is the desntiy plot for the lognormal function; adding values as "f" our impedance function
# OD_test <- OD_test %>%
#   mutate(f_car = dgamma(tt_car, HW_car_motor_gamma$estimate["shape"], HW_car_motor_gamma$estimate["rate"]),
#          f_car = ifelse(is.na(f_car), 0, f_car),
#          f_transit = dlnorm(tt_transit, HW_transit_lnorm$estimate["meanlog"], HW_transit_lnorm$estimate["sdlog"]),
#          f_transit = ifelse(is.na(f_transit), 0, f_transit),
#          f_bike = dlnorm(tt_bike, HW_bike_lnorm$estimate["meanlog"], HW_bike_lnorm$estimate["sdlog"]),
#          f_bike = ifelse(is.na(f_bike), 0, f_bike),
#          f_walk = dlnorm(tt_walk, HW_walk_lnorm$estimate["meanlog"], HW_walk_lnorm$estimate["sdlog"]),
#          f_walk = ifelse(is.na(f_walk), 0, f_walk))

OD_2 <- OD_2 %>%
  mutate(f = case_when(mode == "transit" ~ dlnorm(tt,
                                                  HW_transit_lnorm$estimate["meanlog"], 
                                                  HW_transit_lnorm$estimate["sdlog"]),
                       mode == "car/motor" ~ dgamma(tt,
                                                    HW_car_motor_gamma$estimate["shape"],
                                                    HW_car_motor_gamma$estimate["rate"]),
                       mode == "bike" ~ dlnorm(tt,
                                               HW_bike_lnorm$estimate["meanlog"],
                                               HW_bike_lnorm$estimate["sdlog"]),
                       mode == "walk" ~ dlnorm(tt, 
                                               HW_walk_lnorm$estimate["meanlog"], 
                                               HW_walk_lnorm$estimate["sdlog"])))
```

## Spatial availability

Calculate spatial availabilty:
```{r}
OD_SA <- OD_2 %>%
  mutate(catch = 1) %>%
  mutate(SA_ij = sp_avail(., 
                          o_id = VORIZT1259,
                          d_id = VDESZT1259,
                          pop = workers,
                          opp = jobs,
                          r = catch,
                          f = f))
```

Check that the totals are maintained:
```{r}
sum(OD_SA$SA_ij)
sum(jobs$jobs)
```

Spatial availability by mode:
```{r}
OD_SA %>%
  group_by(mode) %>%
  summarize(V_m = sum(SA_ij))
```

Spatial availability by origin (with check):
```{r}
SA_i <- OD_SA %>%
  group_by(VORIZT1259) %>%
  summarize(V_i = sum(SA_ij))

summary(SA_i)
sum(SA_i$V_i)
```

Spatial availability by origin/mode (with check):
```{r}
SA_im <- OD_SA %>%
  group_by(VORIZT1259, 
           mode) %>%
  summarize(V_im = sum(SA_ij),
            .groups = "drop")

summary(SA_im)
sum(SA_im$V_im)
```

Calculate worker population by zone of origin (with check):
```{r}
population <- workers %>%
  group_by(ZT1259) %>%
  summarize(population = sum(workers))

sum(population$population)
```

Join zonal worker population (with checks):
```{r}
SA_i <- SA_i %>%
  left_join(population,
            by = c("VORIZT1259" = "ZT1259"))
sum(SA_i$V_i)

SA_im <- SA_im %>%
  left_join(population,
            by = c("VORIZT1259" = "ZT1259"))
sum(SA_im$V_im)
```

Calculate availability per capita:
```{r}
SA_i <- SA_i %>%
  mutate(v_i = V_i/population)

summary(SA_i)

SA_im <- SA_im %>%
  mutate(v_im = V_im/population)

summary(SA_im)
```

Join geometry of zones to SA:
```{r}
SA_i <- SA_i %>%
  left_join(zn %>%
              dplyr::select(ZT1259),
            by = c("VORIZT1259" = "ZT1259")) %>%
  st_as_sf()

SA_im <- SA_im %>%
  left_join(zn %>%
              dplyr::select(ZT1259),
            by = c("VORIZT1259" = "ZT1259")) %>%
  st_as_sf()
```

Preliminaries to plotting:
```{r create-bounding-boxes}
#create the bounding box for the Toronto region + a little extra
bbox_new1 <- st_bbox(zn%>%filter(Municipality == "Madrid")) # current bounding box

Madrid_muni_border <- zn %>%
  filter(Municipality == "Madrid") %>%
  st_union()
```

Plot V (spatial availability) by origin:
```{r}
ggplot() +
  geom_sf(data = SA_i,
          aes(fill = V_i),
          color = NA) +
  scale_fill_distiller(palette = "YlGn", #legend scale bar
                       name = expression(v["i"]),
                       na.value = "grey90",
                       direction = 1,
                       trans = "log",
                       breaks=c(0, 10, 100, 1000, 10000))+
  geom_sf(data = Madrid_muni_border, # border for Madrid muni
          colour=alpha("black",1), 
          size = 0.5, 
          fill=NA, 
          show.legend = "polygon") +
  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
                         height = unit(0.8, "cm"), 
                         width = unit(0.8, "cm"),
                         style = north_arrow_orienteering(line_width = 0.25,
                                                          line_col = "dimgrey", 
                                                          fill = c("white","dimgrey"))) +
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
                   height = unit(0.15, "cm")) +
  theme_void() +
  theme(legend.position = c(1.10, 0.25), 
        plot.margin = margin(t = 0, r = 0, b = 0, l = -80)) # positing legend and clipping out white space
```

Plot v (spatial availability per capita) by origin:
```{r}
ggplot() +
  geom_sf(data = SA_i,
          aes(fill = v_i),
          color = NA) +
  scale_fill_gradient2(name = expression(v["i"]),
                       midpoint = 1,
                       na.value = "grey90")+
  geom_sf(data = Madrid_muni_border, # border for Madrid muni
          colour=alpha("black",1), 
          size = 0.5, 
          fill=NA, 
          show.legend = "polygon") +
  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
                         height = unit(0.8, "cm"), 
                         width = unit(0.8, "cm"),
                         style = north_arrow_orienteering(line_width = 0.25,
                                                          line_col = "dimgrey", 
                                                          fill = c("white","dimgrey"))) +
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
                   height = unit(0.15, "cm")) +
  theme_void() +
  theme(legend.position = c(1.10, 0.25), 
        plot.margin = margin(t = 0, r = 0, b = 0, l = -80)) # positing legend and clipping out white space
```

Plot V (spatial availability) by origin-mode:
```{r}
ggplot() +
  geom_sf(data = SA_im,
          aes(fill = V_im),
          color = NA) +
  scale_fill_distiller(palette = "YlGn", #legend scale bar
                       name = expression(v["i"]),
                       na.value = "grey90",
                       direction = 1,
                       trans = "log",
                       breaks=c(0, 10, 100, 1000, 10000))+
  geom_sf(data = Madrid_muni_border, # border for Madrid muni
          colour=alpha("black",1), 
          size = 0.5, 
          fill=NA, 
          show.legend = "polygon") +
  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
                         height = unit(0.8, "cm"), 
                         width = unit(0.8, "cm"),
                         style = north_arrow_orienteering(line_width = 0.25,
                                                          line_col = "dimgrey", 
                                                          fill = c("white","dimgrey"))) +
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
                   height = unit(0.15, "cm")) +
  theme_void() +
  theme(legend.position = c(1.10, 0.25), 
        plot.margin = margin(t = 0, r = 0, b = 0, l = -80)) + # positing legend and clipping out white space
  facet_wrap(~mode, nrow = 2)
```

Plot v (spatial availability per capita) by origin:
```{r}
ggplot() +
  geom_sf(data = SA_im,
          aes(fill = v_im),
          color = NA) +
  scale_fill_gradient2(name = expression(v["i"]),
                       midpoint = 1,
                       na.value = "grey90")+
  geom_sf(data = Madrid_muni_border, # border for Madrid muni
          colour=alpha("black",1), 
          size = 0.5, 
          fill=NA, 
          show.legend = "polygon") +
  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
                         height = unit(0.8, "cm"), 
                         width = unit(0.8, "cm"),
                         style = north_arrow_orienteering(line_width = 0.25,
                                                          line_col = "dimgrey", 
                                                          fill = c("white","dimgrey"))) +
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
                   height = unit(0.15, "cm")) +
  theme_void() +
  theme(legend.position = c(1.10, 0.25), 
        plot.margin = margin(t = 0, r = 0, b = 0, l = -80)) + # positing legend and clipping out white space
  facet_wrap(~mode, nrow = 2)
```
<!--
